\newpage
\addcontentsline{toc}{chapter}{Acknowledgement}
\chapter*{Acknowledgment\markboth{Acknowledgment}{}}

We express our sincere gratitude to the \textbf{Department of Computer Engineering, MIT Academy of Engineering, Alandi (D), Pune} for providing us with the opportunity, academic environment, and resources necessary to undertake our Third Year Major Project, \textit{``SamaySetu --- College Timetable Management System''}.

We extend our heartfelt thanks to our Head of Department, \textbf{Dr. Pramod Ganjewar}, for his vision, guidance, and for providing the departmental infrastructure that enabled us to carry out this work. We are equally grateful to our Project Coordinator, \textbf{Dr. Kanchan Dhote}, for her thoughtful review of our milestones, for structuring the project phases with clarity, and for her constructive feedback at every stage of evaluation.

We wish to express our deepest appreciation to our project guide, \textbf{Dr. Vijaykumar P. Mantri}, whose steady mentorship, technical insight, and patient attention to detail shaped this project from a rough idea into a functional, production-grade system. His willingness to engage with our design decisions, challenge our assumptions, and encourage us to pursue a professional engineering standard has been invaluable. The depth of the system, the quality of its architecture, and the discipline with which we approached each iteration owe a great deal to his guidance.

We also acknowledge the support of all faculty members of the Department of Computer Engineering whose classroom instruction in software engineering, database systems, web technologies, and security fundamentals laid the academic foundation on which this project was built. Their contributions, though indirect, were essential.

We thank our teammates for their collaboration, commitment, and the shared effort that characterised this project. Each member carried a defined set of responsibilities, and the completion of the system in its current form is a product of consistent coordination and mutual respect among the team.

Finally, we express our deepest gratitude to our parents and families for their unwavering moral support, patience, and encouragement throughout our academic journey. Their confidence in us has been the quiet foundation of everything we have accomplished.


\vspace{10mm}

\begin{flushleft}
\begin{tabular}{@{}p{6.0cm}l@{}l@{}}
\textbf{Suryank Dinesh Chavan} \\[1mm]
\textbf{Swarup Patilbuva Daphal} \\[1mm]
\textbf{Abhijeet Purushottam Rane} \\[1mm]
\textbf{Rohan Satishchandra Wani} \\[1mm]
\end{tabular}
\vspace*{3mm}
\end{flushleft}
\vspace*{5mm}

\hrule
